A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-01-20 inst/hankin-AustrianJournalofStatistics.tex
minor tweaks to satisfy AJS style guides

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem has a wide range of applications and I present some
analysis of datasets drawn from the fields of Olympic athletics,
education, and Formula~1 motor racing.

@@

\noindent where $B=1/b$.  Note carefully that these expressions are
likelihoods, not probabilities, and do not sum to 1.  Previous related
work \citep{hankin2024_hyper3} presented the case $n=3$ [in current
  notation] but here arbitrary $n$ may be considered.  Taking the case
$n=8$, for example, gives us $8+1=9$ distinct likelihood functions,
Figure~\ref{ninelikes}.  We see an unanticipated asymmetry between $r$
and $n-r$.  For $r=5$ one might expect that $\hat{a}=\frac{1}{2}$ on
the grounds that the focal competitor's performance was median.
However, $\hat{a}\simeq 0.5886$.  Figure~\ref{dotprobs} shows the
maximum likelihood estimator $\hat{a}$ for various values of $n,r$,
obtained by direct numerical maximization of
equation~\ref{likeforrn3}.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value
  of $r$ indicated in colour by its curve; vertical line shows
  $\hat{a}\simeq 0.5886$ for the case $r=5$}
\end{centering}
\end{figure}

@@

Analysis of expression~\ref{likeforrn3} for large $n$ and $r=pn$ for
fixed $p$ shows that the support
$\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically

@@

Is there evidence that the author's performance differs between years?
Figure~\ref{parkruncontour} shows contours of support [again
  normalised so the support of the evaluate is zero] for the parkrun
dataset, as a function of 2023 strength on the horizontal axis and
2024 strength on the vertical axis.  A null of equal strengths appears
as the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{2023}=0.413$, $\hat{p}_\mathrm{2024}=0.485$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-0.22$, at
${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.441$.  The two units of
support criterion~\citep{edwards1972} is not met, and we conclude that
there is no evidence for a change in competitive strength between the
years.  Now consider table~\ref{fisherparkrun}, giving a summary of
whether nonfocal competitors place ahead of the author, split by year.
A two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
surely an unreasonable conclusion given the author's perceived
performance stagnation.

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.542$, $\hat{p}_\mathrm{applied}=0.852$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.06$, at
${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units of
support criterion~\citep{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.11$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.042$.

@@

%\bibliographystyle{plainat}
\bibliography{hyper2}

>>>>

Here a new simplification of the Plackett--Luce likelihood functions
for order statistics is proposed.  We consider a field of $n$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem has a wide range of applications and I present some
analysis of datasets drawn from the fields of Olympic athletics,
education, and Formula~1 motor racing.

@@

\noindent where $B=1/b$.  Note carefully that these expressions are
likelihoods, not probabilities, and do not sum to 1.  Previous related
work \citep{hankin2024_hyper3} presented the case $n=3$ [in current
  notation] but here arbitrary $n$ may be considered.  Taking the case
$n=9$, for example, gives us 9 distinct likelihood functions,
Figure~\ref{ninelikes}.  We see an unanticipated asymmetry between $r$
and $n-r$.  For $r=5$ one might expect that $\hat{a}=\frac{1}{2}$ on
the grounds that the focal competitor's performance was median.
However, $\hat{a}\simeq 0.5886$.  Figure~\ref{dotprobs} shows the
maximum likelihood estimator $\hat{a}$ for various values of $n,r$,
obtained by direct numerical maximization of
equation~\ref{likeforrn3}.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $r=1,2,\ldots 9$\label{ninelikes}.  Value
  of $r$ indicated in colour by its curve; vertical line shows
  $\hat{a}\simeq 0.5886$ for the case $r=5$}
\end{centering}
\end{figure}

@@

Even though expression ~\ref{likeforrn3} represents a simplification
of the Plackett-Luce likelihood function, it becomes intractable even
for moderate values of $n$ and $r$.  Analysis of
expression~\ref{likeforrn3} for large $n$ and $r=pn$ for fixed $p$
shows that the support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is
asymptotically

@@

Is there evidence that the author's performance differs between years?
Figure~\ref{parkruncontour} shows contours of support [again
  normalised so the support of the evaluate is zero] for the parkrun
dataset, as a function of 2023 strength on the horizontal axis and
2024 strength on the vertical axis.  A null of equal strengths appears
as the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{2023}=0.413$, $\hat{p}_\mathrm{2024}=0.485$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-0.22$, at
${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.441$.  The two units of
support criterion~\citep{edwards1992} is not met, and we conclude that
there is no evidence for a change in competitive strength between the
years.  Now consider table~\ref{fisherparkrun}, giving a summary of
whether nonfocal competitors place ahead of the author, split by year.
A two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
surely an unreasonable conclusion given the author's perceived
performance stagnation.

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.542$, $\hat{p}_\mathrm{applied}=0.852$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.06$, at
${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units of
support criterion~\citep{edwards1992} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.11$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.042$.

@@

%\bibliographystyle{plainat}
\bibliography{simplified.bib}
